---
title: "Introduction to primeDB"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to primeDB}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `primeDB` package provides an R interface to the PRIME (PRIME Database) microbiome data API. This vignette demonstrates how to use the package to access various types of data from the PRIME database, including project metadata, sample information, taxonomic data, abundance data, and statistical analyses.

## Installation and Setup

First, install and load the package:

```{r setup}
# Install the package (if not already installed)
# devtools::install_github("reckonzzz/primeDB")

library(primeDB)
```

## Basic Usage

### Health Check

Before using the API, it's good practice to check if the service is available:

```{r health-check, eval = TRUE}
# Check API health
health <- health_check()
print(health$status)
```

### Database Statistics

Get an overview of the database:

```{r stats, eval = TRUE}
# Get database statistics
stats <- get_stats()
print(stats$basicStats$totalProjects)
print(stats$basicStats$totalSamples)
```

## Working with Projects

### Get All Projects

```{r projects, eval = TRUE}
# Get list of all projects
projects <- get_projects()
print(length(projects))
```

### Get Project Metadata

```{r project-metadata, eval = TRUE}
# Get metadata for specific projects
metadata <- get_project_metadata(c("PRJDB13875", "PRJDB16847"))
head(metadata)
```

### Get Project Samples

```{r project-samples, eval = TRUE}
# Get samples for a specific project
samples <- get_project_samples("PRJDB13875")
print(length(samples))
```

### Get Project Statistics

```{r project-stats, eval = TRUE}
# Get detailed statistics for a specific project
project_stats <- get_project_stats("PRJDB13875")
print(names(project_stats))
```

## Working with Samples

### Get Sample Metadata

```{r sample-metadata, eval = TRUE}
# Get metadata for specific samples
sample_metadata <- get_sample_metadata(c("DRR396974", "DRR396975"))
head(sample_metadata)
```

### Get Sample Statistics

```{r sample-stats, eval = TRUE}
# Get detailed information for a specific sample
sample_stats <- get_sample_stats("DRR396974")
names(sample_stats)
```

### Get Sample Project

```{r sample-project, eval = TRUE}
# Get project information for a specific sample
sample_project <- get_sample_project("DRR396974")
print(sample_project)
```

## Working with Phenotypes

### Get Phenotype List

```{r phenotypes, eval = TRUE}
# Get all phenotypes
phenotypes <- get_phenotypes()
head(phenotypes)
```

### Get Phenotype Projects

```{r phenotype-projects, eval = TRUE}
# Get projects for a specific phenotype
phenotype_projects <- get_phenotype_projects(1)
print(phenotype_projects)
```

### Get Phenotype Samples

```{r phenotype-samples, eval = TRUE}
# Get samples for a specific phenotype
phenotype_samples <- get_phenotype_samples(1)
print(length(phenotype_samples))
```

### Get Phenotype Statistics

```{r phenotype-stats, eval = TRUE}
# Get statistics for a specific phenotype
phenotype_stats <- get_phenotype_stats(1)
names(phenotype_stats)
```

## Working with Body Sites

### Get Body Sites

```{r bodysites, eval = TRUE}
# Get all body sites
bodysites <- get_bodysites()
head(bodysites)
```

### Get Body Site Projects

```{r bodysite-projects, eval = TRUE}
# Get projects for a specific body site
bodysite_projects <- get_bodysite_projects(4)
print(bodysite_projects)
```

### Get Body Site Samples

```{r bodysite-samples, eval = TRUE}
# Get samples for a specific body site
bodysite_samples <- get_bodysite_samples(4)
print(length(bodysite_samples))
```

### Get Body Site Statistics

```{r bodysite-stats, eval = TRUE}
# Get statistics for a specific body site
bodysite_stats <- get_bodysite_stats(4)
names(bodysite_stats)
```

## Working with Countries

### Get Countries

```{r countries, eval = TRUE}
# Get all countries
countries <- get_countries()
head(countries)
```

### Get Country Projects

```{r country-projects, eval = TRUE}
# Get projects for a specific country
country_projects <- get_country_projects(1)
print(country_projects)
```

### Get Country Samples

```{r country-samples, eval = TRUE}
# Get samples for a specific country
country_samples <- get_country_samples(1)
print(length(country_samples))
```

### Get Country Statistics

```{r country-stats, eval = TRUE}
# Get statistics for a specific country
country_stats <- get_country_stats(1)
names(country_stats)
```

## Working with Systems

### Get Systems

```{r systems, eval = TRUE}
# Get all systems
systems <- get_systems()
head(systems)
```

### Get System Projects

```{r system-projects, eval = TRUE}
# Get projects for a specific system
system_projects <- get_system_projects('nervous')
print(system_projects)
```

### Get System Samples

```{r system-samples, eval = TRUE}
# Get samples for a specific system
system_samples <- get_system_samples('nervous')
print(length(system_samples))
```

### Get System Statistics

```{r system-stats, eval = TRUE}
# Get statistics for a specific system
system_stats <- get_system_stats('nervous')
names(system_stats)
```

## Working with Taxonomy

### Get Taxonomy List

```{r taxonomy-list, eval = TRUE}
# Get taxonomy list for a specific rank
taxa <- get_taxonomy_list("gg", "phylum")
head(taxa)
```

### Get Taxonomy Hierarchy

```{r taxonomy-hierarchy, eval = TRUE}
# Get parent taxon information
parent <- get_taxonomy_parent("gg", 15)
print(parent)

# Get child taxa
children <- get_taxonomy_children("gg", 15)
print(children)
```

### Get Taxonomy Statistics

```{r taxonomy-stats, eval = TRUE}
# Get statistics for a specific taxon
taxonomy_stats <- get_taxonomy_stats("gg", 15)
names(taxonomy_stats)
```

### Get Taxonomy Projects

```{r taxonomy-projects, eval = TRUE}
# Get projects for a specific taxon
taxonomy_projects <- get_taxonomy_projects("gg", 15)
print(taxonomy_projects)
```

### Get Taxonomy Samples

```{r taxonomy-samples, eval = TRUE}
# Get samples for a specific taxon
taxonomy_samples <- get_taxonomy_samples("gg", 15)
print(taxonomy_samples)
```

## Working with Abundance Data

### Get Project Abundance Data

```{r abundance-project, eval = TRUE}
# Get abundance data for projects
abundance <- get_abundance_project("gg", "phylum", "absolute", 
                                 c("PRJDB13875", "PRJDB16847"))

# Data is returned as a matrix with samples as rows and taxa as columns
print(dim(abundance))
print(abundance[1:5,1:5])
```

### Get Sample Abundance Data

```{r abundance-sample, eval = TRUE}
# Get abundance data for specific samples
sample_abundance <- get_abundance_sample("gg", "phylum", "absolute", 
                                       c("DRR396974", "DRR396975"))
print(dim(sample_abundance))
print(abundance[1:5,1:5])
```

### Get Cached Abundance Data

```{r abundance-cache, eval = TRUE}
# Get cached abundance data
cached_abundance <- get_abundance_cache("gg", "phylum", "absolute")
print(dim(cached_abundance))
print(abundance[1:5,1:5])
```

## Data Visualization

### Plot Distribution Charts

The package includes a powerful visualization function for creating pie charts from distribution data:

```{r visualization, eval = FALSE}
# Get project statistics
project_stats <- get_project_stats("PRJDB13875")

# Plot all available distributions
plot_distributions(project_stats)

# Plot specific distributions only
plot_distributions(project_stats, plot_names = c("countries", "systems"))
```

The `plot_distributions()` function automatically:
- Creates pie charts for distribution data
- Handles multiple categories with "Others" grouping
- Uses color-coded legends
- Supports combined plots with patchwork
- Provides clear labels with counts and percentages

## Error Handling

The package includes comprehensive error handling:

```{r error-handling, eval = TRUE}
# Handle invalid project accession
tryCatch({
  result <- get_project_samples("INVALID")
}, error = function(e) {
  cat("Error:", e$message, "\n")
})

# Handle invalid database type
tryCatch({
  result <- get_taxonomy_list("invalid", "phylum")
}, error = function(e) {
  cat("Error:", e$message, "\n")
})
```

## Best Practices

1. **Check API health** before making requests using `health_check()`
2. **Handle errors gracefully** using tryCatch
3. **Use appropriate timeouts** for large datasets
4. **Parse CSV data** using readr for abundance data
5. **Cache results** when possible to avoid repeated requests
6. **Use visualization functions** for better data exploration

## Rate Limiting

The API currently has no rate limiting, but it's recommended to:
- Make no more than 10 requests per second
- Implement appropriate delays between requests for large datasets
- Handle errors gracefully

## Package Dependencies

The package requires the following R packages:
- `httr` for HTTP requests
- `jsonlite` for JSON parsing
- `dplyr` for data manipulation
- `readr` for CSV parsing
- `ggplot2` for visualization
- `ggrepel` for label positioning
- `patchwork` for combining plots

## Conclusion

The `primeDB` package provides a comprehensive interface to the PRIME database API. With its easy-to-use functions, comprehensive error handling, and built-in visualization capabilities, it makes accessing and analyzing microbiome data simple and efficient.

For more information, see the package documentation and the [PRIME database website](https://primedb.sjtu.edu.cn). 
